---
execute:
  eval: false
---

# Nominatim US Geocoding Server

## Overview

The lab now maintains a self-hosted [Nominatim](https://nominatim.org/) reverse geocoding server loaded with the full United States OpenStreetMap dataset. This allows for large-scale reverse geocoding (latitude/longitude → address) entirely within the lab's infrastructure, with no data leaving the university network, no rate limits, and no costs.

The server is hosted on the lab's Linux VM and is accessible to anyone on the UW network or connected via [GlobalProtect VPN](globalprotect_mapping.qmd).

**Server:** `http://128.104.50.166/nominatim`

## Accessing the Server

No login or API key is required. You must be either on campus or connected to GlobalProtect VPN. To verify the server is up, navigate to:

```         
http://128.104.50.166/nominatim/status
```

You should see `{"status":0,"message":"OK",...}`.

## Reverse Geocoding in R

### Using `httr`

The `httr` package is the recommended way to query the server. The example below wraps the API call in a function for easy reuse:

```{r}
library(tidyverse)
library(httr)
library(jsonlite)

reverse_geocode_nominatim <- function(lat, lon) {
  response <- GET(
    "http://128.104.50.166/nominatim/reverse",
    query = list(
      lat = lat,
      lon = lon,
      format = "json"
    )
  )

  content(response, as = "parsed")
}
```

Single coordinate:

```{r}
result <- reverse_geocode_nominatim(43.0731, -89.4012)

# Full display name
result$display_name

# Individual address components
result$address$road
result$address$city
result$address$state
result$address$postcode
```

For batch processing a dataframe:

```{r}
coords <- tibble(
  id = 1:3,
  lat = c(43.0731, 41.8781, 44.9778),
  lon = c(-89.4012, -87.6298, -93.2650)
)

geocode_row <- function(lat, lon) {
  result <- reverse_geocode_nominatim(lat, lon)

  tibble(
    display_name = result$display_name %||% NA_character_,
    road         = result$address$road %||% NA_character_,
    city         = result$address$city %||% NA_character_,
    state        = result$address$state %||% NA_character_,
    postcode     = result$address$postcode %||% NA_character_
  )
}

coords |>
  mutate(map2_dfr(lat, lon, geocode_row))
```

## Notes

-   The dataset is the full United States OSM extract, imported April 2026.
-   The server has no rate limiting, but avoid hammering it with highly concurrent requests, as it currently runs on a single CPU.
-   For very large jobs, consider processing in batches and writing intermediate results to disk.
-   If the server appears down, contact Chris (703-715-7091) or feed an LLM the relevant specs of the VM, found [here](virtual_machines.qmd)
